
# Traits inter-dependence

Relationships between traits are important. Examples.

Imagine a simple scenario when there are only two possible relationships between traits (compatible or incompatible), they are simmetric (if A is compatible with B, also B will be compatible with A). Each trait is compatible with itself.

Traits    |  A |  B |  C |  D 
----------|---:|---:|---:|---:
 **A**    | +1 | +1 | -1 | -1 
 **B**    | +1 | +1 | -1 | -1 
 **C**    | -1 | -1 | +1 | +1 
 **D**    | -1 | -1 | +1 | +1    


Have a look at the world

```{r 15.1, message=FALSE}
library(tidyverse)
set.seed(111)
my_world <- matrix(c(1,1,-1,-1,1,1,-1,-1,-1,-1,1,1,-1,-1,1,1), nrow = 4, ncol = 4)
my_world
```

Function for the simulation

```{r 15.2}
traits_inter_dependence <- function(N, t_max, k, mu, p_death, world){
  output <- tibble(trait = as.factor(rep(1:4, each = t_max)), generation = rep(1:t_max, 4), p = rep(NA, t_max * 4))      
  population <- matrix(0, ncol = 4, nrow = N)
  output[output$generation == 1 ,]$p <- colSums(population) / N
  for(t in 2:t_max){
    # innovations
    innovators <- sample(c(TRUE, FALSE), N, prob = c(mu, 1 - mu), replace = TRUE) 
    innovations <- sample(1:4, sum(innovators), replace = TRUE)
    population[cbind(which(innovators == TRUE), innovations)] <- 1
    
    # copying
    demonstrators <- sample(1:N, replace = TRUE)
    demonstrators_traits <- sample(1:4, N, replace = TRUE)

    for(i in 1:N){
      if(population[demonstrators[i], demonstrators_traits[i]]){
        compatibility_score <- sum(world[demonstrators_traits[i], which(population[i,]>0)])
        copy <- (1 / (1 + exp(-k*compatibility_score))) > runif(1)
        if(copy){
          population[i,demonstrators_traits[i]] <- 1
        }
      }
    }
    
    # birth/death
    replace <- sample(c(TRUE, FALSE), N, prob = c(p_death, 1 - p_death), replace = TRUE)
    population[replace, ] <- 0
    
    output[output$generation == t ,]$p <- colSums(population) / N
  }
  output
}
```

```{r 15.3, echo=FALSE }
plot_multiple_traits <- function(data_model) {
  ggplot(data = data_model, aes(y = p, x = generation)) +
    geom_line(aes(colour = trait)) +
    ylim(c(0, 1)) +
    theme_bw() +
    theme(legend.position = "none")
}
```

```{r 15.4}
data_model <- traits_inter_dependence(N = 100, t_max = 1000, k = 10, mu = 0.0005, p_death = 0.01, world = my_world)
plot_multiple_traits(data_model)
```

check that are the compatible traits

```{r 15.5}
data_model[data_model$generation==1000,]
```

It varies at each simulation. What happens by changing the *world*? (remember you can visualise the matrix)

```{r 15.6}
my_world <- matrix(c(1,1,1,-1, 1,1,1,-1,1,1,1,-1,-1,-1,-1,1), nrow = 4, ncol = 4)
data_model <- traits_inter_dependence(N = 100, t_max = 1000, k = 10, mu = 0.0005, p_death = 0.01, world = my_world)
plot_multiple_traits(data_model)
```

All compatible:

```{r 15.7}
my_world <- matrix(c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1), nrow = 4, ncol = 4)
data_model <- traits_inter_dependence(N = 100, t_max = 1000, k = 10, mu = 0.0005, p_death = 0.01, world = my_world)
plot_multiple_traits(data_model)
```

No compatiblity equates to a random copy (remember the logistic)

## Many-traits model


## Summary of the model
  

***

## Analytical appendix

***
  
## Further readings
  

# References
